Natural selection in the Health and Retirement Study

### 2023-12-01 hrs-selection.qmd
Clean up code, get pgi_correct working on DHJ's machine

library(tibble)
library(haven)
library(forcats)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(carData)
library(car, warn.conflicts = FALSE) # masks some
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
 # masks dotchart:
suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))

@@

# TE Maybe not the most efficient method, but this code removes MTAG PGS
pgirep <- select(pgirep,"hhid","pn",matches("PC"),matches("single"))

# TE Going to remove the NEB PGS here
pgirep <- select(pgirep, -c(PGI_NEBwomen_single) )

# TE this creates simple names for PGS in PGI Rep
pgi_rep_names <- str_split_fixed(names(pgirep)[23:57],"_",n=3)[,2]
names(pgirep)[23:57] <- pgi_rep_names

# TE PGI rep does not come pre-standardised. Going to standardize over whole sample to match
# the approach taken with PGENscores
pgirep[,3:ncol(pgirep)] <- scale(pgirep[,3:ncol(pgirep)])

@@

pgirep <- select(pgirep, -matches("^PC\\d"))
#|>
#|
pgen4e_new_names <- left_join(pgen4e_new_names, pgirep, by = join_by(hhid == hhid, pn == pn),suffix=c("","pgirep"))

# TE Height is a duplicated name, but it's with the old PGS, so fixing this manually
colnames(pgen4e_new_names)[which(names(pgen4e_new_names) == "HEIGHT")] <- "HEIGHTpgirep"

@@

pgen4e_new_names <- select(pgen4e_new_names, -hhid, -pn, -any_of(dupe_PGS), -any_of(old_pgs))
pgs_white <- names(pgen4e_new_names)
pgs_white[which(pgs_white == "HEIGHT")] <- "HEIGHTpgirep"

@@

# TE 283 people report having grandkids, but having had no children. Weird!
# This ugly code subsets the data by ethnicity then counts all observations
# where grandkids is greater than 0 but number of kids is 0
gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn == 0, 1, 0)))

gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn == 0, 1, 0)))

# TE if grandkids but no kids div_gkid is infinite. Replacing with NA
rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA

# TE added some variables for grandkid analysis
# TE I've also added sampling weights in calculation of mean values
# note the child weight is just weight* number of kids. so mean_h10gkid 
# and mean_div_h10gkid are actually the same
rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h10gkid = h10gkid/mean_h10gkid,
    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, na.rm = TRUE), 
    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

@@

# TE we can delete more
rm(pgen,pgen4a,pgen4e,pgirep,trk_weights)

@@

Multiple steps were taken to prune the polygenic scores. For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from European ancestry GWAS and GWAS with samples of different ancestry. We choose to use polygenic scores trained only on individuals of European ancestry. we discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.

@@

|               | Low group             | High group            | Difference            |
|------------------|------------------|------------------|------------------|
| Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
| Income        | `r tbl$income_med[1]` | `r tbl$income_med[2]` | `r tbl$income_med[3]` |
| Marriage      | `r tbl$married[1]`    | `r tbl$married[2]`    | `r tbl$married[3]`    |
| Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
| Birth year    | `r tbl$born[1]`       | `r tbl$born[2]`       | `r tbl$born[3]`       |

@@

```{r}
#| label: TE-pgicorrect
#| cache: true

# Writing data for pgi_correct to use
write.table(subset(rand$variables, ethnicity == "white"),"pgi_correct_data.txt",
            quote=FALSE, sep="\t")

@@

# correct_scores holds h2 and R2 values, as well as useful information about the PGS
# such as group categories taken from the PGI repository and simplified names
correct_scores        <- readxl::read_excel("PGI_Heritabilities.xlsx", na = "NA", 
                                            range="B1:U53")

@@

correct_scores[which(correct_scores[, "pgs"] == "CANNABIS"), "pgs"] <- "CANNABISpgirep"
correct_scores[which(correct_scores[, "pgs"] == "DPW"), "pgs"]      <- "DPWpgirep"
correct_scores[which(correct_scores[, "pgs"] == "HEIGHT"), "pgs"]   <- "HEIGHTpgirep"

# some of the PGS have estimates rounded to zero and are thus useless
correct_scores <- subset(correct_scores, chip_h2 > 0)
correct_scores <- subset(correct_scores, R2 > 0)

@@

correct_scores[which(correct_scores$pgs == "AFB"), "twin_h2"]      <- 0.15
correct_scores[which(correct_scores$pgs == "NEBwomen"), "twin_h2"] <- 0.28

@@

# Loop for for calculating results with SNP heritabilities
map_df(1:nrow(correct_scores), function(i) {

system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var', correct_scores[i,"pgs"], ' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2', correct_scores[i,"R2"], ' --h2 ', correct_scores[i,"chip_h2"], ' --weights "weight"', sep=" "))

# This removes "junk" rows in the way of our regression betas
system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
reg_snp <- read.table("reg_results", header=TRUE)

tibble(
  trait = as.character(correct_scores[i,"pgs"]), 
  beta  = as.numeric(reg_snp[1,4]),
  SE    = as.numeric(reg_snp[1,5]),
  Heritability = "Chip"
)
}
) ->
  corrected_effects_snp

# Loop for for calculating results with twin heritabilities
map_df(1:nrow(correct_scores), function(i) {

system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var', correct_scores[i,"pgs"], ' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2', correct_scores[i,"R2"], ' --h2 ', correct_scores[i,"twin_h2"], ' --weights "weight"', sep=" "))

system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
reg_twin <- read.table("reg_results",header=TRUE)

tibble(
  trait = as.character(correct_scores[i,"pgs"]), 
  beta  = as.numeric(reg_twin[1,4]),
  SE    = as.numeric(reg_twin[1,5]),
  Heritability = "Twin"
)
}
) ->
  corrected_effects_twin

corrected_effects           <- rbind(corrected_effects_snp, corrected_effects_twin)
corrected_effects$conf.low  <- corrected_effects$beta - 1.96*corrected_effects$SE
corrected_effects$conf.high <- corrected_effects$beta + 1.96*corrected_effects$SE

@@

corrected_effects <- merge(corrected_effects, correct_scores, 
                           by.x="trait", by.y="Abr",prefix=c("",".y"))

@@

corrected_effects$Category <- case_match(corrected_effects$Category, 
                                        "Health and health behaviours" ~ "Health",
                                        "Personality and well-being" ~ "Personality",
                                        "Fertility and sexual development" ~ "Health",
                                        "Anthropometric" ~ "Health",
                                        "Cognition and education" ~ "Cognition"
                                       )

@@

# This removes ADHD due to the low power of the score. Note ADHD predict RLRS after bonferroni! 
# It also removes NEB
corrected_effects <- subset(corrected_effects, R2 > 0.005)

@@

my_labeller <- function(variable, value) {
  return(value)
  }

corrected_effects |>
  mutate(
    trait = nice_names[trait],
    trait = fct_reorder(trait, beta)
  )|>
  ggplot(aes(y = trait, x = beta, xmin = conf.low, xmax = conf.high,
           color = Heritability)) +
  geom_vline(xintercept = 0, color = "grey60") +
  geom_pointrange(aes(linetype = Heritability == "Twin"),
                  position = position_dodge(0.2)) +
  scale_linetype(guide = "none") +
  coord_cartesian(xlim = c(-0.3, 0.3)) +
  labs(x = "Corrected coefficient on RLRS", y = "") +
  theme_minimal() +
  facet_grid(Category ~ ., scales = "free_y", space = "free_y", labeller = my_labeller)+
  theme(strip.text.y = element_text(angle = -90,face = "bold"))

# Cleaning and saving corrected effects so people can have the exact figures
  corrected_effects$TwinH2_Source <- corrected_effects$Source

  corrected_effects <- corrected_effects[, c("trait", "beta", "SE", "Heritability",
                                            "chip_h2", "twin_h2", "R2", 
                                            "TwinH2_Source", "PoldermanICF_ICD10_subchapter",
                                            "Category")]

 write.csv(corrected_effects,"Corrected_effects.csv")

>>>>

library(tibble)
library(haven)
library(forcats)
library(dplyr, warn.conflicts = FALSE)
library(glue)
library(stringr)
library(purrr)
library(carData)
library(car, warn.conflicts = FALSE) # masks some
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
 # masks dotchart:
suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))

@@

pgirep <- pgirep |> 
  select(hhid, pn, matches("single"), -PGI_NEBwomen_single) |> 
  rename_with(
    \(x) str_replace(x, "PGI_(.*)_single", "\\1")
  )

# TE PGI rep does not come pre-standardised. Going to standardize over whole sample to match
# the approach taken with PGENscores
pgirep[, 3:ncol(pgirep)] <- scale(pgirep[, 3:ncol(pgirep)])

@@

pgen4e_new_names <- left_join(pgen4e_new_names, pgirep, 
                              by = join_by(hhid == hhid, pn == pn),
                              suffix = c("", "pgirep"))

# Height is a duplicated name, but it's with the old PGS, so fixing this manually
pgen4e_new_names <- rename(pgen4e_new_names, 
                           HEIGHTpgirep = HEIGHT)

@@

pgen4e_new_names <- select(pgen4e_new_names, 
                           -hhid, -pn, -any_of(dupe_PGS), -any_of(old_pgs))
pgs_white <- names(pgen4e_new_names)
pgs_white[pgs_white == "HEIGHT"] <- "HEIGHTpgirep"

@@

# 83 people report having grandkids, but having had no children. Weird!
# This code subsets the data by ethnicity then counts all observations
# where grandkids is greater than 0 but number of kids is 0
gkids_unusable_white <- rand |> 
  filter(ethnicity == "white", h10gkid > 0, raevbrn == 0) |> 
  nrow()
gkids_unusable_black <- rand |> 
  filter(ethnicity == "black", h10gkid > 0, raevbrn == 0) |> 
  nrow()

# if grandkids but no kids div_gkid is infinite. Replacing with NA
rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h10gkid = h10gkid/mean_h10gkid,
    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, 
                            na.rm = TRUE), 
    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

@@

# TE we can delete more
rm(pgen, pgen4a, pgen4e, pgirep, trk_weights)

@@

For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from European ancestry GWAS and GWAS with samples of different ancestry. We choose to use polygenic scores trained only on individuals of European ancestry. We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.

@@

|               | Low group             | High group            | Difference            |
|-----------------|-------------------|-------------------|-------------------|
| Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
| Income        | `r tbl$income_med[1]` | `r tbl$income_med[2]` | `r tbl$income_med[3]` |
| Marriage      | `r tbl$married[1]`    | `r tbl$married[2]`    | `r tbl$married[3]`    |
| Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
| Birth year    | `r tbl$born[1]`       | `r tbl$born[2]`       | `r tbl$born[3]`       |

@@

```{r}
#| label: stat-pgicorrect
#| cache: true

# Writing data for pgi_correct to use
readr::write_tsv(
  rand$variables |> filter(ethnicity == "white"),
  "pgi_correct_data.txt",
  quote = "none"
)

@@

# correct_scores holds h2 and R2 values, as well as useful information about the PGS
# such as group categories taken from the PGI repository and simplified names
correct_scores <- readxl::read_excel("PGI_Heritabilities.xlsx", 
                                     na = "NA", range = "B1:U53")

@@

correct_scores$pgs <- case_match(correct_scores$pgs,
  "CANNABIS" ~ "CANNABISpgirep",
  "DPW"      ~ "DPWpgirep",
  "HEIGHT"   ~ "HEIGHTpgirep",
  .default = correct_scores$pgs
)

# some of the PGS have estimates rounded to zero and are thus useless
correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)

@@

correct_scores$twin_h2[correct_scores$pgs == "AFB"] <- 0.15
correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28

@@

run_pgic_once <- function (pgs, r2, h2, h2_type) {
  pgic_command <- glue(
     'python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt', 
     '--outcome "rlrs" --pgi-var {pgs}',
     '--covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A"',
     '"PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2 {r2} --h2 {h2}',
     '--weights "weight"', 
     .sep = " ")
  system(pgic_command)

  # This removes "junk" rows in the way of our regression betas
  system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
  pgic_results <- readr::read_table("reg_results", col_names = TRUE,
                                    col_types = "cdddd")
  file.remove("reg_results")

  tibble(
    trait = pgs,
    beta  = pgic_results[[1, 4]],
    SE    = pgic_results[[1, 5]],
    Heritability = h2_type
  )
}

run_pgic_data <- function (score_data, h2_column, h2_type) {
  results <- map(1:nrow(score_data), 
    \(i) run_pgic_once(pgs = score_data$pgs[i], 
                       r2 = score_data$R2[i],
                       h2 = score_data[i, h2_column, drop = TRUE],
                       h2_type = h2_type)
  )

  list_rbind(results)
}

corrected_effects_snp <- run_pgic_data(correct_scores, h2_column = "chip_h2",
                                       h2_type = "Chip")
corrected_effects_twin <- run_pgic_data(correct_scores, h2_column = "twin_h2",
                                        h2_type = "Twin")

corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
corrected_effects$conf.low <- corrected_effects$beta - 1.96 *
                              corrected_effects$SE
corrected_effects$conf.high <- corrected_effects$beta + 1.96 *
                               corrected_effects$SE

@@

corrected_effects <- left_join(corrected_effects, correct_scores,
                               by = join_by(trait == Abr),
                               suffix = c("", ".y"))

@@

corrected_effects$Category <-
  case_match(
    corrected_effects$Category,
    "Health and health behaviours"     ~ "Health",
    "Personality and well-being"       ~ "Personality",
    "Fertility and sexual development" ~ "Health",
    "Anthropometric"                   ~ "Health",
    "Cognition and education"          ~ "Cognition"
  )

# Cleaning and saving corrected effects so people can have the exact figures
corrected_effects$TwinH2_Source <- corrected_effects$Source
corrected_effects |> 
  select(trait, beta, SE, Heritability, chip_h2, twin_h2, R2, 
           TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
  readr::write_csv("Corrected_effects.csv")

@@

@@

corrected_effects |>
  # This removes ADHD due to the low power of the score. 
  # Note ADHD predict RLRS after bonferroni! 
  # It also removes NEB
  filter(R2 > 0.005) |> 
  mutate(
    trait = nice_names[trait],
    trait = fct_reorder(trait, beta)
  ) |>
  ggplot(aes(y = trait, x = beta, xmin = conf.low, xmax = conf.high,
           color = Heritability)) +
  geom_vline(xintercept = 0, color = "grey60") +
  geom_pointrange(aes(linetype = Heritability == "Twin"),
                  position = position_dodge(0.2)) +
  scale_linetype(guide = "none") +
  coord_cartesian(xlim = c(-0.3, 0.3)) +
  labs(x = "Corrected coefficient on RLRS", y = "") +
  theme_minimal() +
  facet_grid(Category ~ ., scales = "free_y", space = "free_y") +
  theme(strip.text.y = element_text(angle = -90,face = "bold"))
